---
title: "Airport-activity"
author: "Christopher Eshleman"
date: "8/18/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Airport activity: 
## Building a quick profile of daily this and that. 

##### (Code chunk shortcut: Command+Option+I) 

Lemme create a palette. 
Clear my working space, load some useful packages, and set my working drive. 
```{r}
cat("\014") # clear the console
rm(list=ls())
dev.off() 

library(reshape2) 
library(stringr) 
library(plyr)
library(dplyr) 
library(formattable) 
library(doBy) 
library(ggplot2) 

setwd("~/Dropbox/Work and research/Airport-activity/aviation-activity") 
```

Colleagues in the Aviation Department drummed up data on weekday activity for a few random weekdays and weekends across a couple of typical years (2015 and 2016) for activity at the airport I'm looking at. 

The weekday data: 
```{r}
list.files() 
a1 = read.csv("./EWR_FD_150914_150916.csv", skip = 7)
b1 = read.csv("./EWR_FD_160914_160916.csv", skip = 7) 
ab1 = rbind(a1,b1) 
  names(ab1) = tolower(names(ab1)) 
ab1$runway.occupancy.time=NULL
ab1$taxiway.used.to.enter.exit.runway=NULL 
ab1$daytype = "weekday" 
rm(a1,b1) 
```
Locked and loaded. 

The weekend data: 
```{r}
a = read.csv("./EWR_FD_2016_WKND.csv") #read.csv("./EWR_FD_150914_150916.csv", skip = 7)
head(a) 
b = read.csv("./EWR_FD_2015_WKND.csv") #read.csv("./EWR_FD_160914_160916.csv", skip = 7) 
ab = rbind(a,b) 
names(ab) = tolower(names(ab))
ab$daytype = "weekend" 
rm(a,b) 
```

A little messy, so rename variables and then consolidate. 
```{r}
names(ab1) = names(ab) 
ab = rbind(ab1,ab) 
rm(ab1) 
```


## Preparation. 

The airport tracks all kind of activity. But I'm interested explicitly in arrivals and departures. 
```{r}
table(ab$operation) 
ab = subset(ab, ab$operation=="Arrival" | ab$operation == "Departure") 
```

The underlying .csvs and the imports didn't recognize time for what it was. (Times and dates usually require a little extra attention.) 
```{r}
ab$day.time = strptime(ab$event.time, "%m/%d/%y %H:%M", tz = "EST5EDT") 
ab$date = format(as.POSIXct(ab$day.time,format='%m/%d/%Y %H:%M:%S'),format='%m/%d/%Y')
ab$date = as.Date(ab$date, format="%m/%d/%Y") 

ab$weekday = weekdays(ab$date) 
```

Specifically, I want to be able to treat time down to the minute, so I'll just create a time variable that lets me do it easily. 
```{r}
ab$time = as.POSIXct(as.numeric(as.POSIXct(ab$day.time)) %% 86400, origin = "2000-01-01") 
ab$time2 = as.numeric(as.POSIXct(ab$day.time)) %% 86400  # 60*60*24
benchpoints = c("2001-01-01 05:30:00", "2001-01-01 06:30:00", "2001-01-01 9:30:00", "2001-01-01 16:30:00", "2001-01-01 19:30:00", "2001-01-01 23:00:00") 
benchpoints2 = as.numeric(as.POSIXct(benchpoints)) %% 86400 

# Effective hours are everything but 11pm-6am. 86400 seconds in a day. 
ab$timecat = ifelse(ab$time2>=82800 | ab$time2<=21600,"Night","SixAMtoElevenPM") #18000
ab = droplevels(ab) #Quick cleanup to trim unused data levels and 
```

OK, the analysis is going to consider expected delays by time of day. There are some rule-of-thumb level windows of time that the airport uses as categories for high-level analysis. I'll borrow them and shape the data into tables. I'll treat 6 AM to 11 PM as peak and the remaining hours as "night".
```{r}
table(ab$timecat) 
table(ab$timecat,ab$operation) 
ab_wknd = subset(ab,ab$daytype=="weekend") 
ab_wkdy = subset(ab,ab$daytype=="weekday") 
ab_wknd = table(ab_wknd$timecat,ab_wknd$runway) 
ab_wkdy = table(ab_wkdy$timecat,ab_wkdy$runway) 
```


So what DOES activity at the airport look like, anyway? 
```{r}
# Basic density
p = ggplot(ab, aes(x=time2)) + 
  geom_density()
p
# Add mean line
p + geom_vline(aes(xintercept=mean(time2)),
              color="blue", linetype="dashed", size=1)
```

These will be the tables I use to apply expected delays, which we will estimate separately using the same data but during periods of similar delays in other years. 
First I'm curious as to how different the airport's weekday profile is from the weekend: 
```{r}
options(scipen=5)
pq = ggplot(ab, aes(x=time2, fill=daytype)) +
  geom_density(alpha=0.2) + labs(x = "", y = "") 
pq 
ggsave(pq, file="./Activity_EWR.png", height=4,width=6) 
```
The distributions are fairly similar. 

Saving the data: 
```{r}
write.csv(ab_wknd,"./Save weekend 20190818.csv") 
write.csv(ab_wkdy,"./Save weekday 20190818.csv") 
write.csv(ab,"./All runway data 20190818.csv")
```
